Stage-Splitting a Modal Language

### 2014-09-03 paper/semantics.tex
Fix three minor typos.
Addressed Kayvon's comments in section 1-3

Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
subexpressions before any of its stage \bbtwo\ subexpressions. This results in a
stage \bbtwo\ term with no stage \bbone\ content; we say this is a term in a
monostaged
\footnote{In general, \lang could have more than two stages, in which case the
expression would be in a language with one fewer stage.}
language called \langTwo. Then, we perform the remainder of the evaluation with 
$\tworedsym$, an ordinary dynamic semantics for \langTwo.

@@

Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
function:
$\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $v$, via $\reify\xi{v}{f_a}$. By \ref{thm:reify-type}, the resulting
residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.

>>>>

Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
subexpressions before any of its stage \bbtwo\ subexpressions. This results in a
stage \bbtwo\ term with no stage \bbone\ content; we say this is a term in a
monostaged
\footnote{In general, \lang could have more than two stages, in which case the
expression would be in a language with one fewer stage.}
language called \langTwo. Then, we perform the remainder of the evaluation with 
$\tworedsym$, an ordinary dynamic semantics for \langTwo\ (the rules for this
judgment are not shown, but they are standard).

@@

Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
function:
$\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$, which by canonical forms must have the form $v = \next~\hat y$. 
Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $\hat y$, via $\reify\xi{\hat y}{f_a}$. 
Because reification preserves types, the resulting residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.

### 2014-09-03 paper/splitting.tex
Addressed Kayvon's comments in section 1-3

Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
before any arguments are available.
Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
and $f_2$ uses that data structure and the stage-\bbtwo\ input to produce the final output.  
More precisely, a stage splitter is any $s$ such that
\[
	\forall f. \exists f_1,f_2. 
	\left[
		\begin{array}{l}
		s(f) = (f_1,f_2) \text{ and } \\
		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
		\end{array}
	\right]
\]
with the same $\llbracket \cdot \rrbracket$ notation as before.
Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.

As with the dynamics, our stage-splitting algorithm for expressions in \lang\ takes 
the form of two mutually-dependent judgments, $\splitonesym$ and $\splittwosym$. 
Before diving into the full generality, we consider splitting for a simple fragment of the full language: partial values.
After that, we define splitting for full expressions, discuss some notable properties of the algorithm, and then prove correctness.

@@

The basic goal of partial value splitting is to take a partial value $v$ and create from it two single stage values $v_1$ and $v_2$,
where $v_1$ contains only the information from $v$ that is relevant to the first stage, 
and $v_2$ contains only the information from $v$ that is relevant to the second stage.
We call this action {\em masking}.

@@

The distinction between lifting and masking is important because stage \bbone\ expression splitting does both.
To see why, consider the term \verb|(next{hold{1+2} < 5},3*4)|.
The stage \bbone\ result of masking is \verb|((),3*4)|, 
and the precomputation (i.e. the stage \bbone\ result of lifting) is \verb|(1+2)|.
To get the total stage \bbone\ result of splitting, we just take the tuple of those two parts: \verb|(((),3*4),1+2)|.  
This is called the {\em combined result} for stage \bbone.
\footnote{In this example, the combined result is a 2-tuple, but more generally it only must be a term which reduces to a 2-tuple.}

The residual is likewise the combination of the stage \bbtwo\ results of lifting and masking.
In this case, it is ``\verb|fn L => (L<5,())|".

@@

We give the following end-to-end correctness lemmas for open terms. 
It's a bit of a mess currently, but the $\Gamma$ is supposed to be all of the stage \bbone\ bindings, 
whereas $\Gamma'$ is the stage \bbtwo\ bindings.
Substitution splitting works just like value splitting, which is why they use the same symbol.

>>>>

Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
before any arguments are available.
Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
and $f_2$ consumes that data structure and the stage-\bbtwo\ input to produce the final output.  
More precisely, a stage splitter is any $s$ such that
\[
	\forall f. \exists f_1,f_2. 
	\left[
		\begin{array}{l}
		s(f) = (f_1,f_2) \text{ and } \\
		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
		\end{array}
	\right]
\]
with the same $\llbracket \cdot \rrbracket$ notation as before.
Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.

As with the dynamics, our stage-splitting algorithm for expressions in \lang\ takes 
the form of two mutually-dependent judgments, $\splitonesym$ and $\splittwosym$, 
which respectively split terms that type $@ \bbone$ and $@ \bbtwo$. 
Before diving into the full generality, we consider splitting for a simple fragment of the full language: partial values.
After that, we define splitting for full expressions, discuss some notable properties of the algorithm, and then prove correctness.

@@

Recall that, along with a future environment, partial values are the output of the partial evaluation of stage \bbone\ terms.
In addition to base types, tuples, and injections, partial values may also contain references into the future environment.
In this way, partial values express not only stage \bbone\ content, but stage \bbtwo\ content as well.
The goal of partial value splitting is to take a partial value $v$ and create from it two single stage values $v_1$ and $v_2$,
where $v_1$ contains only the information from $v$ that is relevant to first stage code inspecting this value, 
and $v_2$ contains only the information from $v$ that is likewise relevant to second stage code.
We call this action {\em masking}.

@@

The distinction between lifting and masking is important because stage \bbone\ expression splitting does both.
To see why, consider the term \verb|(next{hold{1+2} < 5},3*4)|.
The stage \bbone\ result of masking is \verb|((),3*4)|, 
and the precomputation (i.e. the stage \bbone\ result of lifting) is \verb|(1+2)|.
To get the total stage \bbone\ result of splitting, we just take the tuple of those two parts: \verb|(((),3*4),1+2)|.  
This is called the {\em combined result} for stage \bbone.
\footnote{In this example, the combined result is a 2-tuple, but more generally it only must be a term which reduces to a 2-tuple.}
The splitting residual is likewise the combination of the stage \bbtwo\ results of lifting and masking.
For our example, it is ``\verb|fn L => (L<5,())|".

@@

We give the following end-to-end correctness lemmas for open terms. 
In some of the theorems, we sort the usually heterogeneous context into $\Gamma$ and $\Gamma'$, 
which respectively hold all of the stage \bbone\ and stage \bbtwo\ variable bindings.
Substitution splitting works just like value splitting, which is why they use the same symbol.

### 2014-09-03 paper/techreport.tex
Addressed Kayvon's comments in section 1-3

With these goals in mind, our initial concern is how to express multistage programs. To this end, we have slightly modified prior work to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types. An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.

@@

The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way.  This has previously been explored as an optimization technique for languages with fewer features or a more constrained staging model than \lang. Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.

\Cref{sec:semantics} describes the grammar and type system of \lang, as well as a reference semantics which is essentially a partial evaluator. \Cref{sec:splitting} describes a stage-splitting transformation for \lang\ and proves its equivalence to the reference semantics.
Finally, \Cref{sec:examples} gives several examples of how stage-splitting terms in \lang\ can be used to derive classic data structures, and discusses the effectiveness and limitations of this technique.  Notably, these examples make use of the stage-splitting features not available in prior work.

>>>>

With these goals in mind, our initial concern is how to express multistage programs. To this end, we have modified prior work (\cite{davies96}) to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types. An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.

@@

The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way once inputs are available.  This has previously been explored as an optimization technique for languages with fewer features (notably, no functions) or a more constrained staging model than \lang. Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.  Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.

\Cref{sec:semantics} describes the grammar and type system of \lang, as well as a reference semantics which is essentially a partial evaluator that respects staging. \Cref{sec:splitting} describes a stage-splitting transformation for \lang\ and proves its equivalence to the reference semantics.
Finally, \Cref{sec:examples} gives several examples of how stage-splitting terms in \lang\ can be used to derive classic data structures, and discusses the effectiveness and limitations of this technique.  Notably, these examples make use of the stage-splitting features not available in prior work.
